Application of Turchin's method of statistical regularization

### 2018-03-04 AYSS-2017/article/ayss2017-Zeleniy.tex
Section Prior information completely rewritten

A general case of prior information is assumption about function smoothness. The applications of such prior information is done in three stages:
\begin{enumerate}
\item Minimize additional Shannon's information:
	\begin{equation*}
		I[P(\varphi)] = \int \ln{P(\varphi)} P(\varphi) d\varphi \to min
	\end{equation*}
\item Normalize probability density:
	\begin{equation*}
		\int P(\varphi) d\varphi = 1
	\end{equation*}   
\item Analyze information about smoothness of function $\varphi(x)$:
	\begin{equation}
		\label{eq-smoothness}
		\int \langle \varphi,\hat{\Omega}\varphi \rangle P(\varphi) d\varphi = \omega,
	\end{equation}
where $\omega$ - required level of smoothness,   $\hat{\Omega}$ - operator of smoothness (for example, $\hat{\Omega}=|\frac{d^2}{dx^2}\left\rangle\right\langle\frac{d^2}{dx^2}|$).
\end{enumerate}

The resulting prior probability density for the Gaussian random process:
\begin{equation}
	P_{\alpha}(\vec{\varphi})  = \frac{\alpha^{Rg(\Omega)/2}\det\Omega^{1/2}}{(2\pi)^{N/2}} 
    \exp\left(-\frac{1}{2}(\vec{\varphi},\alpha\Omega\vec{\varphi})\right),
\end{equation}
where  $\alpha = \alpha(\omega)$ - parameter of smoothness.
The prior information depends on $\alpha$-smoothness parameter. This parameter could be estimated in a few different ways: \\
\indent a) select manually using known smoothness, but this is rare probability; \\
\indent b) use most probable parameter: $\alpha^* = \max P(\alpha|f);$ \\
\indent c) use a prior information about smoothness: 
  \begin{equation}\label{eq:varAposteriorAlpha}
      P(\varphi)  = \int P_{\alpha}(\varphi) P(\alpha)~d\alpha ;
  \end{equation}
\indent d) use a posterior information about smoothness: 
  \begin{equation}\label{eq:solveAposteriorAlpha}
      \hat{S}[f] = \int d\alpha \hat{S}_{\alpha}[f] P(\alpha|f).
  \end{equation} 
In fact, methods (c) and (d) are equivalent.

\subsection{Additional prior information}

In eq.~\ref{eq-smoothness} one can use not only smoothness operator, but also different or even multiple different conditions: 
\begin{equation*}
    P_{\alpha\beta}(\varphi) \sim  \det \left|\alpha\Omega_1 +\beta\Omega_2\right|\exp\left(-\frac{1}{2} (\vec{\varphi},(\alpha\Omega_1 + \beta\Omega_2)\vec{\varphi})\right)
\end{equation*}

This features allow to combine different types of parametric prior information. Also one can use non-parametric prior. For example, information about non-negativity:
\begin{equation*}
    P(\varphi) \sim P_{\alpha}(\varphi)P(\varphi(x) > 0)
\end{equation*}

>>>>

We expect that $\varphi$ is relatively smooth and doesn't contain wild
oscillations which naive least squares usually produce. Common way of measuring
function's smoothness is $\langle\varphi,\hat{\Omega}\varphi\rangle$, where
$\hat{\Omega}=|\frac{d^2}{dx^2}\left\rangle\right\langle\frac{d^2}{dx^2}|$.
At the same time we want to minimize information content of prior.

Reasonable choice is ensemble of functions with average smoothness $\omega$ which
maximizes Shannon entropy. Which means we need to maximize:

\begin{equation*}
  H[P(\varphi)] = -\int \ln{P(\varphi)} P(\varphi) d\varphi
\end{equation*}

Subject to constraints:

\begin{equation}
  \int P(\varphi) d\varphi = 1
  \qquad\mbox{and}\qquad
  \int \langle \varphi,\hat{\Omega}\varphi \rangle P(\varphi) d\varphi = \omega,  
  \label{eq-smoothness}
\end{equation}

The resulting prior probability density for the Gaussian random process:
\begin{equation}
  P(\vec{\varphi}|\alpha) = \frac{\det(\alpha\Omega)^{1/2}}{(2\pi)^{N/2}} 
  \exp\left(-\frac{1}{2}\langle\vec{\varphi},\alpha\Omega\vec{\varphi}\rangle\right),
  \label{eq-prior}
\end{equation}

where $\alpha = \alpha(\omega)$~--- parameter of smoothness. Now we have problem
of picking $\alpha$. Information about its value almost never available so we
could consider it random variable as well. This yields us hierarchical Bayes
model:

\begin{equation}
  P(f,\varphi,\alpha) = P(f|\varphi)P(\varphi,\alpha)P(\alpha)
\end{equation}

Since we don't have any information about $\alpha$ we chose hyperprior
$P(\alpha)$ to be flat. Now we have several options to pursue:

\begin{enumerate}
\item Use fully Bayesian approach and marginalize $\alpha$:
  \begin{equation}\label{eq:varAposteriorAlpha}
      P(\varphi) = \int P(\varphi|\alpha) P(\alpha)~d\alpha
  \end{equation}
  In general case MCMC could be used to sample $\varphi$ from full
  posterior. And for Gaussian errors it could be reduced to numerical integral.

\item Use empirical Bayes: pick $\alpha$ corresponding to most probable
  $P(\alpha|f)$

  \begin{equation}
    P(\alpha|f) = \int d\varphi~P(f|\varphi)P(\varphi|\alpha)P(\alpha)
  \end{equation}

  If distribution $P(\alpha|f)$ is narrow it will produce approximately same
  result. And it's the case for our example problem
\end{enumerate}

Derivation above assumed that $\Omega$ if positive definite. It possible to
accommodate singular $\Omega$. Then $\det$ in~(\ref{eq-prior}) should be
replaced with pseudodeterminant. Positive definiteness is desirable since
regularization is guaranteed to work only for non-singular $\Omega$ and singular
or even near-singular $\Omega$ may and likely will cause numerical problems.

Also note that $\Omega$ we use is singular since $\varphi = kx + b$ will have
smoothness equal to zero. One way fix it is to fix value of $\varphi$ to 0 and
boundaries or to put priors on values on
boundaries\cite{calvetti2006aristotelian}. In that case $\alpha\Omega$
in~(\ref{eq-prior}) should be replaced with $\sum_i\alpha_i\Omega_i$ where
different $\Omega_i$ corresponds to smoothness, boundary values etc.

Also it's possible to add any other prior information about function. For
example it possible to add non-negativity:

\begin{equation*}
    P(\varphi) \sim P_{\alpha}(\varphi)P(\varphi(x) > 0)
\end{equation*}
